A weakly structured stem for human origins in Africa

### 2023-01-08 paper.tex
Typos

The rapid rise in sea levels and increased precipitation during MIS
5e interglacial, following a glacial period of aridity across Africa
\citep{Blome2012-lw}, might have triggered migration inland away from the
coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
\citep{Marean2014-pg}. Following these merger events, the stems subsequently fractured into
subpopulations which then persist over the past $\sim$120ka. These
subpopulations can be linked to contemporary groups despite subsequent gene
flow across the continent; for example, a genetic lineage sampled in the Gumuz
has a $0.70$ probability of being inherited from the ancestral ``eastern''
subpopulation $55$ka versus $0.06$ probability of being inherited
from the ``southern'' subpopulation (see Table~\ref{tab:supp-lineage-probabilitys} for
additional comparisons). We also find that Stem 2 continued to contribute to
western Africans during the Last Glacial Period, indicative that this gene flow
likely occurred in western/central Africa (Table~\ref{tab:migration-rates}).
Such an interpretation is reinforced by the differential migration rates
between regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be 5x that
of the rate to eastern Africans during this period. We performed a variety of
validation tests to explore sensitivity of our assumptions, including relaxing
fixed parameters (SI Section~\ref{sec:validation_strategy}). Most validation
tests resulted in parameters similar to the models discussed above. However,
one exception was the inferred Out of Africa and eastern/western African
divergences which were 10-15ka younger than our fixed parameters. Complexity of
thes events warrant the inclusion of additional Eurasian populations and
greater detail in modeling these population expansions; nonetheless, this model
mis-specification does not change other features of our inferred model.

>>>>

The rapid rise in sea levels and increased precipitation during MIS
5e interglacial, following a glacial period of aridity across Africa
\citep{Blome2012-lw}, might have triggered migration inland away from the
coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
\citep{Marean2014-pg}. Following these merger events, the stems subsequently fractured into
subpopulations which then persist over the past $\sim$120ka. These
subpopulations can be linked to contemporary groups despite subsequent gene
flow across the continent; for example, a genetic lineage sampled in the Gumuz
has a $0.70$ probability of being inherited from the ancestral ``eastern''
subpopulation $55$ka versus $0.06$ probability of being inherited
from the ``southern'' subpopulation (see Table~\ref{tab:supp-lineage-probabilitys} for
additional comparisons). We also find that Stem 2 continued to contribute to
western Africans during the Last Glacial Period, indicative that this gene flow
likely occurred in western/central Africa (Table~\ref{tab:migration-rates}).
Such an interpretation is reinforced by the differential migration rates
between regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be 5x that
of the rate to eastern Africans during this period. We performed a variety of
validation tests to explore sensitivity of our assumptions, including relaxing
fixed parameters (SI Section~\ref{sec:validation_strategy}). Most validation
tests resulted in parameters similar to the models discussed above. However,
one exception was the inferred Out of Africa and eastern/western African
divergences which were 10-15ka younger than our fixed parameters. The complexity of
these events warrants the inclusion of additional Eurasian populations and
greater detail in modeling these population expansions; nonetheless, this model
mis-specification does not change other features of our inferred model.
